Textbook Knowledge on Hospital Information Management as Linked Data

### 2019-11-25 paper.tex
Implement or integrate FJs comments.

\begin{abstract}
Textbooks contain abstract knowledge about a domain.
In the case of Hospital Information Management, much of this knowledge consists of systems, their parts and their relations.%, making it a natural fit for RDF, RDFS and OWL.
%As the domain is large and highly relevant to students of Medical Informatics, modelling the knowledge is not only possible but also very useful.
We present a common data model for the domain and the result of applying the data model to three textbooks.
We publish the result over several interfaces that are useful for researchers, practitioners and students, depending on their objectives and their Semantic Web skills.
\end{abstract}

@@

\section{Introduction}
Given a care delivery organization, its health information system (HIS) can be defined as the part of the organization that processes and stores data, information, and knowledge.
It usually consists of a large number of different application systems, computers, and network components~\citep{bb}.
Managing an HIS comprises planning, monitoring and directing activities.
Due to the complexity and the unique conditions in health care, HIS management is an exceptionally challenging task.
There is an enormous amount of frameworks, textbooks and articles describing the scope of HIS management from the perspective of medical informatics.
%However, the disciplines of business informatics and information systems (IS) provide an even broader view on information systems and their management.
A structured representation of the different perspectives leads to a holistic view on HIS management and helps help researchers and students connect their existing knowledge with further knowledge from other sources during research and learning.
In order to integrate different knowledge sources and to provide the knowledge in a structured, machine-readable data format, we extracted knowledge about HIS management from three textbooks~\citep{bb,ob,he} and other sources.
The combination of this knowledge~\citep{einsemantischesnetz,domaene,approachtosupport} results in SNIK, the Semantic Network of Information Management in Hospitals (\enquote{Krankenhaus} in German).

In order to encourage and enable other researchers, students and health informatics professionals to use available knowledge of HIS management, we answer the following questions in this paper:

We discuss advantages and disadvantages of the interfaces and give recommendations on which ones present the best compromise for different use cases and target audiences.
We conclude with plans for future work on interlinking and visualization.

As the domain is large and highly relevant to students of Medical Informatics, modelling the knowledge is not only possible but also very useful.

@@

Using a data model of Information Management 
Depending on the learning style and information need of the student, several tools are available to cosnume that knowledge.

@@

The Meta Model (see \cref{fig:metamodel}) provides a common vocabulary for the domain of HIS management and thus defines, which superclasses and relations can be used.
SNIK version 0.4.1 comprises five subontologies that are built upon the Meta Model, see Table 1.
At the head of the class hierarchy is the \enquote{Top} class, which has exactly three disjunctive subclasses.
Following the Meta Model, each class has to be a subclass of exactly one of them.
The correct superclass of a new concept can be found by answering the question: Who (\enquote{Role}) does what (\enquote{Function}) and which information (\enquote{EntityType}) is needed? If a concept is neither of them, it cannot be modeled using the Meta Model.
As the subclass relation is transitive, a new class can be placed further down the hierarchy and it can still be inferred, whether it is a Function, Role or EntityType (see Figure 3).
Besides the subclass relationship, two classes can be connected with relations provided by the Meta Model.
The generic \enquote{is associated with} relationship is possible between all classes but carries little information.
Which more specific relations between two classes are available depends on their superclasses.
For example, a role and a function can be connected as \enquote{is involved in} \enquote{is responsible for} and \enquote{approves}.
Relations that are neither of them can either be modeled by using the generic \enquote{is associated with} relation or by creating and using a new sub relation of \enquote{is associated with}

\section{Concessions to Practicality}
Our main focus is that SNIK provides a net benefit to its users, many of which do not have Semantic Web expertise.
This led to several concessions to usability and actual usage by sacrificing full conformance to the W3C OWL recommendation in the following cases:

\paragraph{OWL Restrictions}
The meta model offers several properties between the different subtopclasses.
The definitions of \aurl{rdfs}{domain} and \aurl{rdfs}{range} state that the subjects, respectively objects of all triples with that property are instances of that class.
Because SNIK contains classes, which are subclasses, not instances of the subtopclasses, using those properties directly in SNIK is technically wrong and OWL property restrictions should be used.
This presented the following problems:
\begin{itemize}
\item Tools such as LodLive don't understand those OWL constructs and only show the RDF form with blank nodes and helper constructs, which is confusing to the users
\item Communicating OWL restrictions to extractors and non-Semantic Web expert project members is difficult.
\item Extractors only enter the name of the property, not whether it is a "someValuesFrom" or "allValuesFrom" restriction.
For example, it is unclear whether each CEO is involved in at least one project review, or whether each CEO is only involved in project reviews.
It may even be the case that both of those statements are not what the author intended.
\item Deleting or adding a connection between classes in the OntoWiki is no longer an atomic operation and may result in unreferenced blank nodes or partially restored undo results.
\end{itemize}

@@

\begin{figure*}
\caption{The SNIK Meta Model}
\label{fig:metamodel}
\includegraphics[width=0.8\textwidth]{img/metamodel9s.pdf}
\end{figure*}

@@

\section{Data Model}\label{sec:architecture}
\citet{domaene} contains an initial structure of the meta model.
\todo{Describe the meta model here}
The \textbf{Semantic Network of Information Management in Hospitals} (SNIK\footnote{Hospital means \enquote{Krankenhaus} in German.}) is a modular OWL 2 DL ontology.
The \enquote{Meta Model} defines three basic disjunctive classes and their possible relations: Roles (who), Function (does what) and Entity Types (and which information is therefore needed).
A set of modular subontologies define subclasses of those three classes and their relations as described by a certain knowledge source about information management in hospitals:
%from different sources:% three textbooks, an interview and a standard.

@@

After the initial publish phase, we describe the repeating steps of the reuse phrase with the SNIK lifecycle in \Cref{fig:lifecycle}:
\begin{enumerate}
\item the Virtuoso SPARQL endpoint is used for querying and as data source for all our applications  
\item an instance of the OntoWiki is used for punctual changes and additions by non-Semantic-Web-experts, supplemented by SPARUL for changes involving large numbers of classes 
\item LIMES~\citep{limes} generates interlink candidates between the different textbooks, which can be approved or rejected by the users
\item LodView\footnote{\url{https://lodview.it/}} lets the user browse among the classes of SNIK and view particular classes in detail
\item SNIK Graph, detailed in \Cref{sec:snik-graph}, supplements the RDF browser by visualizing the relationships between classes.
Users often discover incorrect modelling or possibilities for enhancement in SNIK graph.
\item SNIK quality is a web application that uses SPARQL queries to find problems of varying degree, such as violations of domain, range or naming conventions, subclass cycles or missing definitions.
\item Problems suggested by users and SNIK quality are verified, solved, and published on the SPARQL endpoint using SPARUL queries, at which point the circle begins anew. 
\end{enumerate}

@@

\section{Applications}
SNIK was initially intended~\citep{domaene} both as the data source for software to support hospital CIOs and to support teaching.
For the former goal, two applications were designed and implemented as prototypes~\citep{toreonto}: (1) the requirements engineering decision support system TOREOnto and (2) the knowledge exploration and navigation visualization CIONx.
Due to internal regulations, efforts to integrate them into the informational infrastructure of the Uniklinikum Leipzig were stopped.
Consequently, the other approaches are in the area of teaching support:

@@

\begin{figure}
\caption{Students learn new concepts about HIS quality by linking them to concepts already learned.
A teacher asks a student to find out how the new concept \enquote{Quality of Data} is linked to the \enquote{Patient Identification Number}.
The student connects the two concepts by using the “spiderworm“ visualization and learns that a patient identification number is associated with object identity.
Object identity is a subclass of integrity of data.
Besides integrity of data, there are also 13 other criteria for quality of data.}
\label{fig:snik-graph-spiderworm}
\includegraphics[width=\columnwidth]{img/snik-graph-spiderworm.png}
\end{figure}

SNIK Graph is a web application\footnote{\url{http://www.snik.eu/graph}} that transforms the classes and triples of SNIK to nodes and edges of a graph.
The graph is visualized using the Cytoscape.js~\citep{cytoscape} library with the force-directed Euler layout, see~\cref{fig:snik-graph-overview}.
With several thousand classes, specific parts of SNIK can be hard to discern.
Thus, there are several options to view subgraphs of SNIK, for example to show only a specific chapter of a book to prepare a lecture about a specific topic.
Users can also search for and restrict the view to a single class and then show the neighbourhood of that class (see \cref{fig:snik-graph-circle-star}) and subsequently the neighbourhood of selected neighbours of the previous step.
SNIK Graph can also calculate the shortest path to between classes.
We also tried to automatically find the most interesting path but this was not successfull as it is subjective and the depends on the goal of the user.
Path and neighourhood operations are joined in the \enquote{spider worm} (see \cref{fig:snik-graph-spiderworm}), which consists of the shortest path between a start node and an end node together with the end node’s neighbourhood, illustrate the context of a concept.

@@

%The technical environment of SNIK is characterized in previous work [11].
\section{Discussion}
SNIK is based on the Meta Model from Figure \todo{reference}.
Thus, SNIK can be considered as an archetype for ontologies describing for a given domain (here \enquote{HIS management}), what functions a certain role has to carry out and what information a person of this role needs and what information she/he provides while carrying out the functions.
This may be used for domains like the medical department of a hospital as well.
This would be valuable knowledge, for instance as basis for systems analysis projects [1].
Different user groups can use the interfaces of SNIK that are most suitable to them.
For example, the IT management of a hospital can use SNIK as a vocabulary to integrate data from different formats that result from application components from different vendors.
Using the SPARQL1 query language, experts can also integrate applications with the SNIK ontology.
 Experts can also use the SPARQL query editor to answer specific questions, see Table 2.
Students and teachers, on the other hand, benefit more from SNIK Graph, which intuitively presents knowledge without the need for technical skills.

@@

\section{Future Work}
We plan to extend the text-based interlinks with semantic-similarity-based ontology matching [5] to find more correspondencies between the textbooks. This will enable users to switch to another textbook when a certain concept is explained there in more detail. The SNIK toolset, SNIK Graph in particular, is under permanent improvement. For instance, we are considering alternatives to the shortest path that are the most preferred by different audiences and whether those paths can be (semi-)automatically generated using suitable node or edge weight functions. We are also investigating candidates for additional subontologies and as interlink targets.

\section{Acknowledgments}
This work is supported by the DFG (German Research Foundation) under the Projects SNIK, grant no. 1605/7-1 and 1387/8-1, as well as HITO, grant no. WI 1605/11-1 and I3726-N31.
%We thank the publishers ... and ... for granting us permission to publish the subontologies based on the textbooks \citet{bb,ob,he} under an open license.
%\nocite{*} 
\bibliographystyle{ios1}
\bibliography{paper,snik}

\section{Notes to Reviewers}
This paper includes and extends text and figures from:
- SNIK Graph: \todo{include medinfo poster reference}

>>>>

\begin{abstract}
Textbooks contain abstract knowledge about a domain.
In the case of Hospital Information Management, much of this knowledge consists of information management functions, roles executing these functions  and the information used or updated by these functions.
%, making it a natural fit for RDF, RDFS and OWL.
%As the domain is large and highly relevant to students of Medical Informatics, modelling the knowledge is not only possible but also very useful.
We present a common data model for the domain and the result of applying the data model to three textbooks an interview and a standard.
We publish the result over several interfaces that are useful for researchers, practitioners and students, depending on their objectives and their Semantic Web skills.
\end{abstract}

@@

\section{Introduction}
Given a care delivery organization, its health information system (HIS) can be defined as the part of the organization that processes and stores data, information, and knowledge.
It usually consists of different application systems, computers, and network components~\citep{bb}.
Managing an HIS comprises planning, monitoring and directing activities.
Due to the complexity and the unique conditions in health care, HIS management is an exceptionally challenging task.
There is an enormous amount of frameworks, textbooks and articles describing the scope of HIS management from the perspective of medical informatics.
%However, the disciplines of business informatics and information systems (IS) provide an even broader view on information systems and their management.
A structured representation of the different perspectives leads to a holistic view on HIS management and helps help researchers and students connect their existing knowledge with further knowledge from other sources during research and learning.
In order to integrate different knowledge sources and to provide the knowledge in a structured, machine-readable data format, we extracted knowledge about HIS management from three textbooks~\citep{bb,ob,he} and other sources.
The combination of this knowledge~\citep{einsemantischesnetz,domaene,approachtosupport} results in the SNIK ontology, the Semantic Network of Information Management in Hospitals (\enquote{Krankenhaus} in German).

In order to encourage and enable other researchers, students and health informatics professionals to use available knowledge of HIS management, we answer the following questions in this paper:
\todo{add questions}
We introduce different user interfaces to the SNIK ontology and discuss their suitability for different use cases of the target audiences.
%We discuss advantages and disadvantages of the interfaces and give recommendations on which ones present the best compromise for different use cases and target audiences.
We conclude with plans for future work on interlinking and visualization.

%As the domain is large and highly relevant to students of Medical Informatics, modelling the knowledge is not only possible but also very useful.
Medical informatics students, who are trained for executive positions in information management departments of healthcare institutions, such as hospitals, need a clear terminology of their domain.
Due to different frameworks and textbooks dealing with information management in healthcare, modelling the knowledge unravels the links between the different views on information management.
These are only implicitly known or not known at all by experts in the field.

@@

Using a data model of Information Management 
Depending on the learning style and information need of the student, several tools are available to consume that knowledge.
\todo{FJ: Welchen Zweck hat dieser Teil in der Einleitung? Auch würde ich die Ziele vielleicht etwas kompakter als Liste beschreiben.}

@@

The meta model (see \cref{fig:metamodel}) provides a common vocabulary for the domain of HIS management and thus defines, which superclasses and property can be used.
SNIK version 0.4.1 comprises five subontologies that are built upon the meta model, see Table 1.
At the head of the class hierarchy is the \enquote{Top} class, which has exactly three disjunctive subclasses.
Following the meta model, each class has to be a subclass of exactly one of them.
The correct superclass of a new concept can be found by answering the question: Who (\enquote{Role}) does what (\enquote{Function}) and which information (\enquote{EntityType}) is needed? If a concept is neither of them, it cannot be modeled using the meta model.
As the subclass relation is transitive, a new class can be placed further down the hierarchy and it can still be inferred, whether it is a Function, Role or EntityType (see Figure 3).
Besides the subclass relationship, two classes can be connected with relations provided by the meta model.
The generic \enquote{is associated with} relationship carries little information.
For example, a role and a function can be connected as \enquote{is involved in} \enquote{is responsible for} and \enquote{approves}.
Relations that are neither of them can either be modeled by using the generic \enquote{is associated with} relation or by creating and using a new sub relation of \enquote{is associated with}

\section{Concessions to Practicality}
SNIK is mainly used by students and researchers having no or little semantic web knowledge.
This led to several concessions to usability and actual usage by sacrificing full conformance to the W3C OWL recommendation in the following cases:

\paragraph{OWL Restrictions}
The meta model offers several properties between the different subtopclasses.
The definitions of \aurl{rdfs}{domain} and \aurl{rdfs}{range} state that the subjects, respectively objects of all triples with that property are instances of that class.
The classes of SNIK are subclasses, not instances, of the subtopclasses.
This means that using those properties directly in SNIK is technically wrong and OWL property restrictions should be used.
This presented the following problems:
\begin{itemize}
\item Tools such as LodLive don't understand those OWL constructs and only show the RDF form with blank nodes and helper constructs, which is confusing to the users
\item Communicating OWL restrictions to extractors and non-Semantic Web expert project members is difficult.
\item Extractors only enter the name of the property, not whether it is a "someValuesFrom" or "allValuesFrom" restriction.
For example, it is unclear whether each CEO is involved in at least one project review, or whether each CEO is only involved in project reviews.
It may even be the case that both of those statements are not what the author intended.
\todo{FJ: Sollte man hier evtl. noch erklären, wie es denn tatsächlich gemeint ist? Eigentlich haben die in SNIK modellierten Aussagen jeweils eine Bedeutung wie: Laut Lehrbuch xy muss/sollte ein CEO im Krankenhaus i. d. R. an Projektreviews (innerhalb von IM-Projekten) beteiligt sein. Ob ein Krankenhaus einen oder oder mehrere CEOs hat, ist uns an der Stelle egal, ebenso wie viele Projektreviews das sind. Wir drücken hier sozusagen eher eine Regel (laut Lehrbuch) aus.  }
\item Deleting or adding a connection between classes in the OntoWiki\todo{needs to be introduced first} is no longer an atomic operation and may result in unreferenced blank nodes or partially restored undo results.
\end{itemize}

@@

\begin{figure*}
\caption{The SNIK meta model}
\label{fig:metamodel}
\includegraphics[width=0.8\textwidth]{img/metamodel9s.pdf}
\end{figure*}

@@

\section{Data Model}\label{sec:architecture}
\citet{domaene} contains an initial structure of the meta model.
\todo{Describe the meta model here}
The \textbf{Semantic Network of Information Management in Hospitals} (SNIK\footnote{Hospital means \enquote{Krankenhaus} in German.}) is a modular OWL 2 DL ontology.
The \enquote{meta model} defines three basic disjunctive classes and their possible relations: Roles (who), Function (does what) and Entity Types (and which information is therefore needed).
A set of modular subontologies define subclasses of those three classes and their relations as described by a certain knowledge source about information management in hospitals:
%from different sources:% three textbooks, an interview and a standard.

@@

After the initial publish phase, we describe the repeating steps of the reuse phrase with the SNIK lifecycle in \Cref{fig:lifecycle}:
\begin{enumerate}
\item the Virtuoso SPARQL endpoint is used for querying and as data source for all our applications  
\item an instance of the OntoWiki~\citep{ontowiki} is used for punctual changes and additions by non-Semantic-Web-experts, supplemented by SPARUL for changes involving large numbers of classes 
\item LIMES~\citep{limes} generates interlink candidates between the different textbooks, which can be approved or rejected by the users
\item LodView\footnote{\url{https://lodview.it/}} lets the user browse among the classes of SNIK and view particular classes in detail
\item SNIK Graph, detailed in \Cref{sec:snik-graph}, supplements the RDF browser by visualizing the relationships between classes.
Users often discover incorrect modelling or possibilities for enhancement in SNIK graph.
\item SNIK quality is a web application that uses SPARQL queries to find problems of varying degree, such as violations of domain, range or naming conventions, subclass cycles or missing definitions.
\item Problems suggested by users and SNIK quality are verified, solved, and published on the SPARQL endpoint using SPARUL queries, at which point the circle begins anew. 
\end{enumerate}

@@

\section{Applications}
SNIK was initially intended~\citep{domaene} both as  software to support hospital CIOs and to support teaching.
For the former goal, two applications were designed and implemented as prototypes~\citep{toreonto}: (1) the requirements engineering decision support system TOREOnto and (2) the knowledge exploration and navigation visualization CIONx.
Due to internal regulations, efforts to integrate them into the informational infrastructure of the Uniklinikum Leipzig were stopped.
Consequently, the other approaches are in the area of teaching support:

@@

\begin{figure}
\caption{Students learn new concepts about HIS quality by linking them to concepts already learned.
A teacher asks a student to find out how the new concept \enquote{Quality of Data} is linked to the \enquote{Patient Identification Number}.
The student connects the two concepts by using the “spiderworm“ visualization and learns that a patient identification number is associated with object identity.
Object identity is a subclass of integrity of data.
Besides integrity of data, there are also 13 other criteria for quality of data.}
\label{fig:snik-graph-spiderworm}
\includegraphics[width=\columnwidth]{img/snik-graph-spiderworm.png}
\end{figure}
\todo{FJ: In diesem Abschnitt fehlt die Diskussion, wie es durch die verschiedenen Zielgruppen bei Use Cases genutzt werden kann.}
SNIK Graph is a web application\footnote{\url{http://www.snik.eu/graph}} that transforms the classes and triples of SNIK to nodes and edges of a graph.
The graph is visualized using the Cytoscape.js~\citep{cytoscape} library with the force-directed Euler layout, see~\cref{fig:snik-graph-overview}.
With several thousand classes, specific parts of SNIK can be hard to discern.
Thus, there are several options to view subgraphs of SNIK, for example to show only a specific chapter of a book to prepare a lecture about a specific topic.
Users can also search for and restrict the view to a single class and then show the neighbourhood of that class (see \cref{fig:snik-graph-circle-star}) and subsequently the neighbourhood of selected neighbours of the previous step.
SNIK Graph can also calculate the shortest path to between classes.
We also tried to automatically find the most interesting path but this was not successfull as it is subjective and the depends on the goal of the user.
Path and neighourhood operations are joined in the \enquote{spider worm} (see \cref{fig:snik-graph-spiderworm}), which consists of the shortest path between a start node and an end node together with the end node’s neighbourhood, illustrate the context of a concept.

@@

%The technical environment of SNIK is characterized in previous work [11].
\section{Discussion}
SNIK is based on the meta model from Figure \todo{reference}.
Thus, SNIK can be considered as an archetype for ontologies describing for a given domain (here \enquote{HIS management}), what functions a certain role has to carry out and what information a person of this role needs and what information she/he provides while carrying out the functions.
This may be used for domains like the medical department of a hospital as well.\todo{FJ: Ich glaube, das müsste man ausführlicher erklären, das würde ich eher ans Ende der Diskussion nehmen.}
This would be valuable knowledge, for instance as basis for systems analysis projects [1].
Different user groups can use the interfaces of SNIK that are most suitable to them.
For example, the IT management of a hospital can use SNIK as a vocabulary to integrate data from different formats that result from application components from different vendors.
Using the SPARQL1 query language, experts can also integrate applications with the SNIK ontology.
 Experts can also use the SPARQL query editor to answer specific questions, see Table 2.
Students and teachers, on the other hand, benefit more from SNIK Graph, which intuitively presents knowledge without the need for technical skills.

@@

\section{Future Work}
\todo{Connection with HITO}
We plan to extend the text-based interlinks with semantic-similarity-based ontology matching [5] to find more correspondencies between the textbooks. This will enable users to switch to another textbook when a certain concept is explained there in more detail. The SNIK toolset, SNIK Graph in particular, is under permanent improvement. For instance, we are considering alternatives to the shortest path that are the most preferred by different audiences and whether those paths can be (semi-)automatically generated using suitable node or edge weight functions. We are also investigating candidates for additional subontologies and as interlink targets.

\section{Acknowledgments}
This work is supported by the DFG (German Research Foundation) under the Projects SNIK, grant no. 1605/7-1 and 1387/8-1, as well as HITO, grant no. WI 1605/11-1 and I3726-N31.
%We thank the publishers ... and ... for granting us permission to publish the subontologies based on the textbooks \citet{bb,ob,he} under an open license.
%\nocite{*} 
\todo{FJ: Hmmm, ziemlich selbstreferenziell. Brauchen wir da wirklich so viele Publikationen von uns? Welche anderen könnte man noch einbringen? Ich schaue auch mal, was mir so über den Weg läuft.}
\bibliographystyle{ios1}
\bibliography{paper,snik}
\section{Notes to Reviewers}
This paper includes and extends text and figures from:
- SNIK Graph: \todo{include medinfo poster reference}
